---
title: CCA / two-block correlation inference with multifer
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{CCA / two-block correlation inference with multifer} %\VignetteEngine{knitr::knitr} %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette walks the **canonical correlation** case: inference on
the ordered correlations between paired blocks `X` and `Y`. This is a
shipped part of `multifer`'s latent-root surface, with an explicit
support boundary:

- the exact whitened-space stepwise deflation is implemented for the
  paired-row design, common-`Z` nuisance-adjusted designs, and
  nuisance-adjusted designs with within-block exchangeability
- richer structured designs fall back to the conservative first-root
  cap until their validity story is fully settled

## Setup

```{r setup}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(multifer)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

## A paired-row CCA example

```{r synthetic-cca}
set.seed(1)
n   <- 60
p_x <- 6
p_y <- 5
k   <- 2
U   <- qr.Q(qr(matrix(rnorm(n * k), n, k)))
Ax  <- matrix(rnorm(p_x * k), p_x, k)
Ay  <- matrix(rnorm(p_y * k), p_y, k)
# Correlated latent factor with noise on each block
X <- U %*% t(Ax) + matrix(rnorm(n * p_x, sd = 0.4), n, p_x)
Y <- U %*% t(Ay) + matrix(rnorm(n * p_y, sd = 0.4), n, p_y)
```

## Running `infer_cca()`

The default public wrapper uses the dedicated `cancor_cross` adapter,
which claims only `(cross, correlation)` and therefore has no
strict-dispatch ambiguity:

```{r infer-cca}
res <- infer_cca(
  X, Y,
  design = paired_rows(),
  B = 99,
  R = 20,
  alpha = 0.05,
  seed = 2
)
res
```

## Running `infer()` with the `cross_svd` reference adapter

The `cross_svd` adapter claims **both** `(cross, covariance)` and
`(cross, correlation)` capabilities. Under strict dispatch, that means
you must spell out `relation` — calling `infer(geometry = "cross", ...)`
alone is an error because the relation is ambiguous.

```{r infer-cross-svd}
res <- infer(
  adapter  = "cross_svd",
  data     = list(X = X, Y = Y),
  geometry = "cross",
  relation = "correlation",
  design   = paired_rows(),
  B        = 99,
  R        = 20,
  alpha    = 0.05,
  seed     = 2
)
res
```

## Using the `cancor_cross` adapter explicitly

If you want to spell out the correlation-only adapter directly,
`cancor_cross` wraps `stats::cancor()` and claims only
`(cross, correlation)`:

```{r infer-cancor}
res2 <- infer(
  adapter  = "cancor_cross",
  data     = list(X = X, Y = Y),
  geometry = "cross",
  relation = "correlation",
  B        = 99,
  R        = 20,
  alpha    = 0.05,
  seed     = 3
)
res2
```

## What the shipped CCA boundary means in practice

1. First-root inference is always valid under the paired-row design.
2. Stepwise multi-root inference is valid when the engine can project
   both blocks into an exchangeability-preserving residual basis —
   which is done automatically for `paired_rows()`, for
   `nuisance_adjusted(Z)`, and for `nuisance_adjusted(Z, groups)` with
   grouped exchangeability.
3. For a design not yet in the exact multi-root list,
   such as `exchangeable_rows()`, `run_cross_ladder()` caps `max_steps`
   at 1 automatically. The
   first-root rejection is still valid; the package simply refuses
   to make a stepwise claim it cannot yet back up.

The rank-deficient failure mode has an executable check:

```{r validity, error = TRUE}
bad_X <- X
bad_X[, p_x] <- 2 * bad_X[, 1L] + 0.5 * bad_X[, 2L]  # exact collinearity
infer(adapter = "cross_svd", data = list(X = bad_X, Y = Y),
      geometry = "cross", relation = "correlation",
      B = 49, R = 4, seed = 4)
```

## Reading subspace stability first

When canonical correlations are close, the canonical pairs are only
identifiable up to rotation. `auto_subspace = TRUE` (the default)
bundles tied roots into a subspace unit and `print()` leads with the
principal-angle stability.

```{r subspace-cca}
head(res$subspace_stability)
```
