---
title: CCA / two-block correlation inference with multifer
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{CCA / two-block correlation inference with multifer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 6,
  fig.height = 4,
  dpi       = 150
)
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(family = params$family,
                                              preset = params$preset))
}
```

```{r albers-classes, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family, params$preset
))
```

```{r library}
library(multifer)
```

Use `infer_cca()` when each row of `X` and `Y` is the same
observational unit and the question is whether successive canonical
correlations survive a paired-row null. This vignette walks that
canonical-correlation case and then shows where the shipped support
boundary matters.

## A paired-row CCA example

```{r synthetic-cca}
set.seed(1)
n   <- 60
p_x <- 6
p_y <- 5
k   <- 2
U   <- qr.Q(qr(matrix(rnorm(n * k), n, k)))
Ax  <- matrix(rnorm(p_x * k), p_x, k)
Ay  <- matrix(rnorm(p_y * k), p_y, k)
# Correlated latent factor with noise on each block
X <- U %*% t(Ax) + matrix(rnorm(n * p_x, sd = 0.4), n, p_x)
Y <- U %*% t(Ay) + matrix(rnorm(n * p_y, sd = 0.4), n, p_y)
```

## Running `infer_cca()`

The default public wrapper uses the dedicated `cancor_cross` adapter,
which claims only `(cross, correlation)` and therefore has no
strict-dispatch ambiguity:

```{r infer-cca}
res <- infer_cca(
  X, Y,
  design = paired_rows(),
  B = 99,
  R = 20,
  alpha = 0.05,
  seed = 2
)
res
```

```{r check-cca-wrapper, include = FALSE}
stopifnot(
  is_infer_result(res),
  identical(res$provenance$adapter_id, "cancor_cross"),
  nrow(res$component_tests) >= 1L,
  all(is.finite(res$component_tests$p_value)),
  all(res$component_tests$p_value >= 0 & res$component_tests$p_value <= 1)
)
```

## Running `infer()` with the `cross_svd` reference adapter

The `cross_svd` adapter claims **both** `(cross, covariance)` and
`(cross, correlation)` capabilities. Under strict dispatch, that means
you must spell out `relation` — calling `infer(geometry = "cross", ...)`
alone is an error because the relation is ambiguous.

```{r infer-cross-svd}
res <- infer(
  adapter  = "cross_svd",
  data     = list(X = X, Y = Y),
  geometry = "cross",
  relation = "correlation",
  design   = paired_rows(),
  B        = 99,
  R        = 20,
  alpha    = 0.05,
  seed     = 2
)
res
```

```{r check-cca-cross-svd, include = FALSE}
stopifnot(
  is_infer_result(res),
  identical(res$provenance$adapter_id, "cross_svd"),
  nrow(res$component_tests) >= 1L,
  all(is.finite(res$component_tests$p_value))
)
```

## Using the `multivarious_cca` backend

`multivarious_cca` wraps `multivarious::cca()` and is available as an
explicit backend. Its component-test scaffold is parity-pinned against
`cancor_cross` on the shipped CCA support matrix, including paired rows,
blocked rows, nuisance-adjusted designs, grouped nuisance-adjusted
designs, first-root capping outside the multi-root boundary, and
rank-deficient validity failures.

It is still not the `infer_cca()` default. The remaining maturity work
is adapter-owned: loading estimates, score estimates, score intervals,
subspace stability summaries, regularization defaults, and future
feature-evidence statistics are normalization-specific rather than
raw-equal across backends.

```{r infer-multivarious-cca}
mv_res <- infer(
  adapter  = "multivarious_cca",
  data     = list(X = X, Y = Y),
  geometry = "cross",
  relation = "correlation",
  design   = paired_rows(),
  B        = 99,
  R        = 0,
  alpha    = 0.05,
  seed     = 2
)
mv_res$component_tests
```

```{r check-multivarious-cca, include = FALSE}
stopifnot(
  is_infer_result(mv_res),
  identical(mv_res$provenance$adapter_id, "multivarious_cca"),
  isTRUE(all.equal(
    mv_res$component_tests$statistic,
    res$component_tests$statistic,
    tolerance = 1e-10
  )),
  isTRUE(all.equal(
    mv_res$component_tests$p_value,
    res$component_tests$p_value,
    tolerance = 1e-8
  ))
)
```

## What this vignette tests

```{r cca-labels, echo=FALSE, results='asis'}
cat(knitr::kable(
  data.frame(
    piece = c("Maturity", "Statistic", "Null", "Estimand"),
    value = c(
      "mature",
      res$mc$statistic_label,
      res$mc$null_label,
      res$mc$estimand_label
    ),
    check.names = FALSE
  ),
  col.names = c("piece", "value")
), sep = "\n")
```

The statistic, null, and estimand values above are live labels from
`res$mc`; the text below explains the support boundary around those
engine claims.

## Using the `cancor_cross` adapter explicitly

If you want to spell out the correlation-only adapter directly,
`cancor_cross` wraps `stats::cancor()` and claims only
`(cross, correlation)`:

```{r infer-cancor}
res2 <- infer(
  adapter  = "cancor_cross",
  data     = list(X = X, Y = Y),
  geometry = "cross",
  relation = "correlation",
  B        = 99,
  R        = 20,
  alpha    = 0.05,
  seed     = 3
)
res2
```

## What the shipped CCA boundary means in practice

1. First-root inference is always valid under the paired-row design.
2. Stepwise multi-root inference is valid when the engine can project
   both blocks into an exchangeability-preserving residual basis —
   which is done automatically for `paired_rows()`, for
   `nuisance_adjusted(Z)`, and for `nuisance_adjusted(Z, groups)` with
   grouped exchangeability.
3. For a design not yet in the exact multi-root list,
   such as `exchangeable_rows()`, `run_cross_ladder()` caps `max_steps`
   at 1 automatically. The
   first-root rejection is still valid; the package simply refuses
   to make a stepwise claim it cannot yet back up.

The rank-deficient failure mode has an executable check:

```{r validity, error = TRUE}
bad_X <- X
bad_X[, p_x] <- 2 * bad_X[, 1L] + 0.5 * bad_X[, 2L]  # exact collinearity
infer(adapter = "cross_svd", data = list(X = bad_X, Y = Y),
      geometry = "cross", relation = "correlation",
      B = 49, R = 4, seed = 4)
```

## Reading subspace stability first

When canonical correlations are close, the canonical pairs are only
identifiable up to rotation. `auto_subspace = TRUE` (the default)
bundles tied roots into a subspace unit and `print()` leads with the
principal-angle stability.

The same boundary applies at the feature level. Signed canonical
coefficients or structure correlations can be summarized as bootstrap
ratios for identifiable singleton roots, but those ratios are evidence
summaries rather than p-values. Null-calibrated feature p-values should
target an invariant statistic, such as squared coefficient, squared
structure correlation, or a subspace norm, so sign flips and rotations
do not change the claim.

```{r subspace-cca}
head(res$subspace_stability)
```
