---
title: Choosing designs and null actions in multifer
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Choosing designs and null actions in multifer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 6,
  fig.height = 4,
  dpi       = 150
)
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(family = params$family,
                                              preset = params$preset))
}
```

```{r albers-classes, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family, params$preset
))
```

```{r library}
library(multifer)
```

You choose a `multifer` design when the rows of your data carry
structure that changes which perturbations are valid. For a two-block
analysis, the most common case is simple: row `i` of `X` and row `i` of
`Y` describe the same observation, so the null must break the
cross-block pairing without changing either block's marginal geometry.

The design is the third part of a typed shape:

```{r typed-shape}
typed_shape(
  geometry("cross"),
  relation("covariance"),
  paired_rows()
)
```

That object says what data shape you have, what root family you are
testing, and which row perturbations the engine may use.

## What does a paired-row design do?

Here is a small two-block covariance example. The first latent score is
shared by `X` and `Y`, so the leading covariance root should stand out
from paired-row noise.

```{r make-cross-data}
set.seed(1)
n <- 48
p_x <- 7
p_y <- 5

score <- scale(rnorm(n), center = TRUE, scale = FALSE)
X <- matrix(rnorm(n * p_x, sd = 0.5), n, p_x)
Y <- matrix(rnorm(n * p_y, sd = 0.5), n, p_y)
X[, 1:3] <- X[, 1:3] + score %*% matrix(c(1.4, -1.1, 0.9), 1)
Y[, 1:2] <- Y[, 1:2] + score %*% matrix(c(1.2, -0.8), 1)
```

```{r paired-plsc}
res <- infer_plsc(
  X, Y,
  design = paired_rows(),
  B = 49,
  R = 8,
  seed = 4
)
res$component_tests
```

```{r check-paired-plsc, include = FALSE}
stopifnot(
  is_infer_result(res),
  nrow(res$component_tests) >= 1L,
  all(is.finite(res$component_tests$p_value)),
  all(res$component_tests$p_value >= 0 & res$component_tests$p_value <= 1),
  isTRUE(res$units$selected[[1L]]),
  res$component_tests$p_value[[1L]] <= 0.05
)
```

For this shape, `paired_rows()` tells `infer_plsc()` to use the
cross-block null action appropriate for matched observations. If you
omit the design, the wrapper uses this same paired-row default for
cross-block inference.

## Which design matches your rows?

Use the design that matches the exchangeability claim you would be
willing to defend.

```{r design-examples}
paired_rows()

groups <- rep(c("site_a", "site_b", "site_c"), each = 16)
blocked_rows(groups)

Z <- cbind(intercept = 1, age = seq(-1, 1, length.out = n))
nuisance_adjusted(Z)
```

```{r check-design-examples, include = FALSE}
stopifnot(
  is_design(paired_rows()),
  is_design(blocked_rows(groups)),
  is_design(nuisance_adjusted(Z)),
  length(blocked_rows(groups)$groups) == n,
  nrow(nuisance_adjusted(Z)$Z) == n
)
```

`paired_rows()` is for matched cross-block rows. `blocked_rows(groups)`
keeps permutations inside known exchangeability blocks. Use
`nuisance_adjusted(Z)` when the null should be conditional on nuisance
covariates, with one row of `Z` for each observation.

## What does the design protect?

The design is not metadata for display. It is part of the validity
contract. If the rows no longer match, a cross-block call should fail
before the engine spends time fitting ladders or bootstraps.

```{r row-mismatch, error = TRUE}
infer_plsc(
  X[-1, ], Y,
  design = paired_rows(),
  B = 19,
  R = 4,
  seed = 5
)
```

The same principle applies to richer designs. A blocked design with
missing block labels is not a valid exchangeability statement:

```{r bad-blocks, error = TRUE}
bad_groups <- groups
bad_groups[1] <- NA
blocked_rows(bad_groups)
```

## How does CCA differ?

CCA uses the same design vocabulary, but the correlation relation has a
stricter stepwise validity boundary. The public `infer_cca()` wrapper
uses the dedicated `cancor_cross` adapter and supports paired rows, a
common nuisance matrix, and nuisance-adjusted designs with within-block
exchangeability.

```{r cca-design}
cca_res <- infer_cca(
  X, Y,
  design = paired_rows(),
  B = 39,
  R = 6,
  seed = 6
)
head(cca_res$component_tests)
```

```{r check-cca-design, include = FALSE}
stopifnot(
  is_infer_result(cca_res),
  identical(cca_res$provenance$adapter_id, "cancor_cross"),
  nrow(cca_res$component_tests) >= 1L,
  all(is.finite(cca_res$component_tests$p_value)),
  all(cca_res$component_tests$p_value >= 0 & cca_res$component_tests$p_value <= 1)
)
```

If a design is outside the exact multi-root CCA boundary, the engine
keeps the first-root claim conservative rather than silently extending
the stepwise ladder.

## Where to go next

- `vignette("pca-inference")` for one-block variance inference.
- `vignette("plsc-inference")` for paired two-block covariance.
- `vignette("cca-inference")` for the CCA support boundary.
- `vignette("writing-adapters")` for declaring adapter capabilities and
  checked assumptions.
