---
title: LDA / generalized-eigen inference with multifer
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{LDA / generalized-eigen inference with multifer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 6,
  fig.height = 4,
  dpi       = 150
)
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(family = params$family,
                                              preset = params$preset))
}
```

```{r albers-classes-hidden, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family, params$preset
))
```

```{r library}
library(multifer)
```

This vignette walks the **generalized-eigen** flagship in `multifer`:
linear discriminant analysis (LDA). The object of inference is the
ordered sequence of discriminant roots, not specific variables and not
specific class-vs-class contrasts.

In typed-shape language, LDA lives at:

- geometry: `"geneig"`
- relation: `"generalized_eigen"`
- design: exchangeable labels under the null

The mathematical engine is described in `notes/engine_geneig_spec.md`.
That note is the canonical reference for the whitening map, the
B-metric deflation rule, and the admissibility constraints that make the
geneig family valid.
The narrower public-wrapper contract is recorded in
`notes/lda_maturity_contract.md`.

The public geneig surface is intentionally narrow in v1: `infer_lda()`
ships discriminant-root significance for LDA, while broader
metric-weighted / contrastive generalized-eigen families remain future
extensions rather than hidden features of the current wrapper.
The default `lda_refit` adapter is also a classical full-rank LDA path:
because it delegates fitting to `MASS::lda()`, p > n - K and
collinear within-class designs are rejected as validity failures rather
than stabilized silently.

## What the null actually tests

For LDA, the null action is **label permutation**. The rows of `X` stay
in place; only the class labels are permuted. That is the right null for
the question "is there discriminant structure beyond noise?" because it
destroys the class assignment while preserving the marginal geometry of
the observations themselves.

This is different from a row-permutation null. Row permutation would
change the sample geometry, which is not the claim LDA is making. The
geneig engine therefore delegates the null action to the adapter, and
the `lda_refit` adapter uses label permutation only.

## Why the ladder is naturally short

With `K` classes, classical LDA has at most `K - 1` non-zero
discriminant directions. In finite samples, the returned unit table may
still include trailing numerical roots from the fitted object, but the
scientific question remains the short discriminant-root ladder rather
than an open-ended component sequence.

## A planted discriminant example

We build a three-class problem where the classes differ strongly along
one axis and only weakly elsewhere. The example is intentionally small:
it gives you a compact result object to inspect without turning the
vignette into a calibration study.

```{r lda-sim}
set.seed(1)
n_per_class <- 24
labels <- factor(rep(c("a", "b", "c"), each = n_per_class))
means <- rbind(
  c(-4, 0, 0, 0),
  c( 0, 0, 0, 0),
  c( 4, 0, 0, 0)
)

X <- do.call(rbind, lapply(seq_len(nrow(means)), function(i) {
  matrix(rnorm(n_per_class * ncol(means), sd = 0.4), ncol = ncol(means)) +
    matrix(means[i, ], nrow = n_per_class, ncol = ncol(means), byrow = TRUE)
}))
```

## Running `infer_lda()`

The public wrapper defaults to `adapter = "lda_refit"` and to
`targets = "component_significance"`. At the moment this wrapper is
about discriminant-root significance; the public bootstrap stability
path for the geneig family is not exposed yet.

```{r infer-lda}
res <- infer_lda(
  X,
  labels,
  B = 39,
  alpha = 0.05,
  seed = 41
)
res
```

```{r check-lda-result, include = FALSE}
stopifnot(
  is_infer_result(res),
  identical(res$provenance$adapter_id, "lda_refit"),
  nrow(res$component_tests) >= 1L,
  all(is.finite(res$component_tests$p_value)),
  all(res$component_tests$p_value >= 0 & res$component_tests$p_value <= 1)
)
```

## What this vignette tests

```{r lda-labels, echo=FALSE, results='asis'}
cat(knitr::kable(
  data.frame(
    piece = c("Maturity", "Statistic", "Null", "Estimand"),
    value = c(
      "narrow",
      res$mc$statistic_label,
      res$mc$null_label,
      res$mc$estimand_label
    ),
    check.names = FALSE
  ),
  col.names = c("piece", "value")
), sep = "\n")
```

The statistic, null, and estimand values above are read from `res$mc`
so this vignette reports the engine's current provenance labels.

```{r lda-units}
res$units
res$component_tests
```

This example is meant to show the structure of the returned
`infer_result`. On stronger class separation or with a larger Monte
Carlo budget, one or more discriminant roots may clear the
label-permutation null; on this lightweight vignette budget the example
may instead show a conservative no-selection outcome.

## Reading the result correctly

The result is about **roots**. A significant first root does **not**
mean:

- that any specific variable is individually significant
- that any specific pair of classes is individually significant
- that the loadings are p-values

Those are different inferential objects. `infer_lda()` itself does
not claim `variable_significance` — that target is gated as a sidecar
API rather than a frozen sub-block of `infer_result`.

## A sidecar API for variable-level p-values

For the question "which variables drive the discriminant axes?",
`feature_importance_pvalues()` runs a null-calibrated permutation
test on aggregate squared-loading importance, using the adapter's
own label-permutation null and a `refit_data` hook that translates
the geneig operator payload back to LDA `(X, y)` data.

For generalized-eigen methods, raw Euclidean loadings are rarely the
right default estimand. A mature LDA adapter should expose a
metric-aware feature statistic, such as a B-metric discriminant
contribution or class-separation contribution, through the
feature-evidence hook family. The loading-based sidecar below is a
small worked example of the calibration route, not a claim that raw
loadings are the final LDA feature statistic.

```{r lda-importance}
adapter <- adapter_lda_refit()
data    <- list(X = X, y = labels)
recipe  <- infer_recipe(
  geometry = "geneig", relation = "generalized_eigen",
  adapter  = adapter,  targets  = "component_significance"
)
fit     <- adapter$refit(NULL, data)
rank    <- ncol(adapter$loadings(fit))
units   <- form_units(adapter$roots(fit)[seq_len(rank)])

fi <- feature_importance_pvalues(
  recipe, adapter, data = data, units = units,
  original_fit = fit, B = 49, adjust = "BH", seed = 11
)
fi[, c("variable", "statistic", "p_value", "p_adjusted")]
```

```{r check-lda-importance, include = FALSE}
stopifnot(
  nrow(fi) == ncol(X),
  all(is.finite(fi$statistic)),
  all(fi$p_value >= 0 & fi$p_value <= 1),
  which.min(fi$p_adjusted) == 1L
)
```

```{r lda-scatter, echo = FALSE, fig.cap = "Discriminant scores colored by class. The first axis separates the planted class means; the second shows residual sample-level structure."}
fit <- adapter$refit(NULL, data)
S <- adapter$scores(fit)
op <- par(mar = c(4, 4, 3, 1))
tryCatch({
  plot(
    S[, 1L], if (ncol(S) >= 2L) S[, 2L] else rep(0, nrow(S)),
    col = c("#1f77b4", "#2ca02c", "#d62728")[as.integer(labels)],
    pch = 19L, xlab = "LD1", ylab = "LD2",
    main = "LDA discriminant scores"
  )
  legend("topright", legend = levels(labels), bty = "n",
         col = c("#1f77b4", "#2ca02c", "#d62728"), pch = 19L)
}, finally = par(op))
```

Variable 1 -- the planted discriminant axis -- carries the lowest
adjusted p-value. The result is a
companion to `res$component_tests`, not a replacement: roots and
variables are different inferential objects.

## Why deflation is B-metric, not Euclidean

The generalized-eigen problem is:

`A v = lambda B v`

with `B` symmetric positive definite. After testing the leading root,
the remaining ladder steps must deflate in the **B-metric**. Euclidean
deflation gives the wrong remaining generalized roots. That is why the
registration-time gate requires geneig adapters to mark their
`residualize` hook as B-metric-aware (or to declare that the engine
itself owns the deflation).

The derivation and the regression tests live in
`notes/engine_geneig_spec.md` and `tests/testthat/test-engine-geneig.R`.

## Scope boundary

This vignette is deliberately narrow:

- it covers LDA as the current flagship `(geneig, generalized_eigen)`
  family
- it does not claim contrastive PCA or metric-weighted PCA are already
  shipped
- it does not promise variable-level significance

Those remain separate extensions, not hidden features of the current
public surface.
