---
title: LDA / generalized-eigen inference with multifer
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{LDA / generalized-eigen inference with multifer} %\VignetteEngine{knitr::knitr} %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette walks the **generalized-eigen** flagship in `multifer`:
linear discriminant analysis (LDA). The object of inference is the
ordered sequence of discriminant roots, not specific variables and not
specific class-vs-class contrasts.

In typed-shape language, LDA lives at:

- geometry: `"geneig"`
- relation: `"generalized_eigen"`
- design: exchangeable labels under the null

The mathematical engine is described in `notes/engine_geneig_spec.md`.
That note is the canonical reference for the whitening map, the
B-metric deflation rule, and the admissibility constraints that make the
geneig family valid.

The public geneig surface is intentionally narrow in v1: `infer_lda()`
ships discriminant-root significance for LDA, while broader
metric-weighted / contrastive generalized-eigen families remain future
extensions rather than hidden features of the current wrapper.

## Setup

```{r setup}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(multifer)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

## What the null actually tests

For LDA, the null action is **label permutation**. The rows of `X` stay
in place; only the class labels are permuted. That is the right null for
the question "is there discriminant structure beyond noise?" because it
destroys the class assignment while preserving the marginal geometry of
the observations themselves.

This is different from a row-permutation null. Row permutation would
change the sample geometry, which is not the claim LDA is making. The
geneig engine therefore delegates the null action to the adapter, and
the `lda_refit` adapter uses label permutation only.

## Why the ladder is naturally short

With `K` classes, classical LDA has at most `K - 1` non-zero
discriminant roots. The sequential ladder inherits that bound: there is
no reason to test beyond the maximum discriminant rank the model can
support.

## A planted rank-1 discriminant example

We build a three-class problem where the classes differ strongly along
one axis and only weakly elsewhere. The planted signal has one dominant
discriminant direction, so the inferential target is:

- root 1: signal
- root 2: noise

```{r lda-sim}
set.seed(1)
n_per_class <- 24
labels <- factor(rep(c("a", "b", "c"), each = n_per_class))
means <- rbind(
  c(-4, 0, 0, 0),
  c( 0, 0, 0, 0),
  c( 4, 0, 0, 0)
)

X <- do.call(rbind, lapply(seq_len(nrow(means)), function(i) {
  matrix(rnorm(n_per_class * ncol(means), sd = 0.4), ncol = ncol(means)) +
    matrix(means[i, ], nrow = n_per_class, ncol = ncol(means), byrow = TRUE)
}))
```

## Running `infer_lda()`

The public wrapper defaults to `adapter = "lda_refit"` and to
`targets = "component_significance"`. At the moment this wrapper is
about discriminant-root significance; the public bootstrap stability
path for the geneig family is not exposed yet.

```{r infer-lda}
res <- infer_lda(
  X,
  labels,
  B = 39,
  alpha = 0.05,
  seed = 41
)
res
```

```{r lda-units}
res$units
res$component_tests
```

This example is meant to show the structure of the returned
`infer_result`. On stronger class separation or with a larger Monte
Carlo budget, one or more discriminant roots may clear the
label-permutation null; on this lightweight vignette budget the example
may instead show a conservative no-selection outcome.

## Reading the result correctly

The result is about **roots**. A significant first root does **not**
mean:

- that any specific variable is individually significant
- that any specific pair of classes is individually significant
- that the loadings are p-values

Those are different inferential objects. The current LDA/geneig surface
does not claim `variable_significance`, and this vignette does not
reinterpret stability summaries as significance tests.

## Why deflation is B-metric, not Euclidean

The generalized-eigen problem is:

`A v = lambda B v`

with `B` symmetric positive definite. After testing the leading root,
the remaining ladder steps must deflate in the **B-metric**. Euclidean
deflation gives the wrong remaining generalized roots. That is why the
registration-time gate requires geneig adapters to mark their
`residualize` hook as B-metric-aware (or to declare that the engine
itself owns the deflation).

The derivation and the regression tests live in
`notes/engine_geneig_spec.md` and `tests/testthat/test-engine-geneig.R`.

## Scope boundary

This vignette is deliberately narrow:

- it covers LDA as the current flagship `(geneig, generalized_eigen)`
  family
- it does not claim contrastive PCA or metric-weighted PCA are already
  shipped
- it does not promise variable-level significance

Those remain separate extensions, not hidden features of the current
public surface.
