---
title: Getting started with multifer
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Getting started with multifer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 6,
  fig.height = 4,
  dpi       = 150
)
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(family = params$family,
                                              preset = params$preset))
}
```

```{r albers-classes, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family, params$preset
))
```

```{r library}
library(multifer)
```

## The problem

You ran a PCA on a 40 × 8 data matrix. It gave you eight singular values,
ordered from largest to smallest. You now need to decide how many of those
eight components carry real latent structure and how many are just noise
fitting fluctuations.

That decision is harder than it looks. Random-matrix eigenvalues have their
own non-trivial distribution, so no fixed cutoff works. After you remove the
leading component, the null distribution of the *next* root depends on the
one you just removed. And when two true roots are close, the individual
loadings aren't identifiable at all — only the subspace they span is.

`multifer` is a **typed perturbation inference layer** for that question.
It is not a fitting framework — you fit your model in `multivarious`
(or any other package that provides the required accessors via a thin
adapter) and `multifer` provides the inference layer on top: the exact
collapsed Vitale P3 ladder under sequential deflation, automatic
subspace detection when roots are near-tied, and executable validity
contracts that fail fast on bad input. The division of labor is
deliberate: `multifer` should not be in the business of reinventing
PCA / PLSC / CCA fitters, and `multivarious` already ships a tested
projector hierarchy.

## First win: one line of inference

Here is a 40 × 8 matrix with three true signal components planted on
top of Gaussian noise:

```{r make-data}
dat <- bench_oneblock_shadowing(
  n = 40, p = 8,
  root_profile = c(6, 3, 1.5),
  seed = 1
)
dim(dat$X)
dat$meta$true_rank
```

`infer_pca()` is the convenience wrapper. It picks the
`multivarious_pca` adapter by default (falling back to
`prcomp_oneblock` if `multivarious` isn't installed) and runs the
full ladder + bootstrap stability sweep:

```{r infer-pca}
res <- infer_pca(dat$X, B = 99, R = 20, seed = 2)
res
```

Read the header: the ladder tested the latent ordering, rejected
through rung 3, and froze 3 selected component units. That is the
answer — three components are statistically detectable at the chosen
α. Everything else is in the result object, ready for downstream use.

### Reusing a pre-fit projector

If you've already fit a projector and want to keep that exact fit
(for prediction, projection, or other downstream work), pass it
through `model = `:

```{r infer-with-model, eval = FALSE}
fit <- multivarious::pca(dat$X)
res <- infer_pca(dat$X, model = fit, B = 99, R = 20, seed = 2)
```

This is equivalent to the wrapper-only call above; it just skips the
internal refit on the original data.

## The mental model: typed shapes

`multifer` organizes every inferential question as a **typed shape** — a
triple of `(geometry, relation, design)`:

| Piece | What it fixes | Examples |
|:---|:---|:---|
| `geometry` | what data shape we operate on | `oneblock`, `cross` |
| `relation` | which operator whose roots we test | `variance`, `covariance`, `correlation` |
| `design` | which null action is valid for this data | `paired_rows()`, `blocked_rows(g)`, `nuisance_adjusted(Z)` |

PCA is `(oneblock, variance, trivial)`. PLSC is
`(cross, covariance, paired_rows)`. CCA is
`(cross, correlation, paired_rows)` routed through a whitened-space
stepwise deflation. The triple determines what the engine is allowed to
claim — if you pair an adapter with an ambiguous relation under strict
dispatch (the default), `multifer` errors instead of silently guessing.

Most users never call `infer()` directly. Five thin wrappers map each
mature or narrow-surface family to a one-line call:

```{r wrappers, eval = FALSE}
infer_pca (X)             # oneblock variance        (mature)
infer_plsc(X, Y)          # cross covariance         (mature)
infer_cca (X, Y)          # cross correlation        (mature)
infer_lda (X, labels)     # geneig discriminant roots (narrow)
infer_plsr(X, Y)          # predictive-gain PLSR      (narrow)
```

The lower-level `infer()` stays available when you need explicit control
over the adapter, the geometry, or a non-default design. The default
`infer_cca()` path uses the dedicated `cancor_cross` adapter so the
public CCA wrapper stays single-relation and unambiguous.

## Reading the result

An `infer_result` is a frozen unit-centered object. Six schema blocks
answer six distinct questions:

```{r result-structure}
names(res)
```

The two you touch most often are `units` and `component_tests`:

```{r units}
res$units
```

```{r component-tests}
res$component_tests
```

`units` says which latent objects exist and which passed the ladder. Each
row is either a single component or a bundled subspace (more on that
below). `component_tests` carries one row per tested rung: the observed
test statistic, the sequential p-value, the Monte Carlo standard error,
and the exceedance count `r`.

Three stability blocks give you bootstrap-based reliability summaries,
keyed by `unit_id`:

```{r stability}
head(res$variable_stability)
head(res$subspace_stability)
```

`variable_stability` is a *stability measure*, not a p-value — the schema
separates those two kinds of claims explicitly. `subspace_stability`
carries mean and max principal angles between the reference fit and each
bootstrap replicate, which is the right object when loadings are only
defined up to rotation.

### Visualizing the ladder decisions

The p-value sequence across rungs plus the α threshold makes the ladder's
decision visible at a glance:

```{r ladder-plot, echo = FALSE, fig.cap = "Component p-values and the 0.05 threshold. Green bars were rejected (declared significant), red bars were not."}
pvals <- res$component_tests$p_value
units <- res$component_tests$unit_id
col   <- ifelse(pvals <= 0.05, "#2ca02c", "#d62728")
bp <- barplot(
  pvals,
  names.arg = units,
  col       = col,
  border    = NA,
  ylim      = c(0, max(0.12, max(pvals) * 1.2)),
  ylab      = "sequential p-value",
  xlab      = "unit",
  main      = "multifer ladder decisions"
)
abline(h = 0.05, lty = 2, col = "gray30")
text(x = bp, y = pvals + 0.005,
     labels = sprintf("%.3f", pvals), cex = 0.8)
```

The plot is just the `component_tests$p_value` column against its
unit id — everything the decision rests on is already in the result.

## When roots are near-tied: subspace inference

When two true roots are nearly equal, their individual loading directions
aren't identifiable. What *is* identifiable is the 2-dimensional subspace
they span. `auto_subspace = TRUE` (the default) detects near-ties
automatically and reports principal-angle stability on the bundled
subspace as the headline.

```{r tied-data}
set.seed(13)
n <- 80; p <- 6; k <- 3
U <- qr.Q(qr(matrix(rnorm(n * k), n, k)))
V <- qr.Q(qr(matrix(rnorm(p * k), p, k)))
sig <- c(8.0, 8.0, 3.0)
X_tied <- U %*% diag(sig * sqrt(n - 1), nrow = k) %*% t(V) +
  matrix(rnorm(n * p, sd = 0.01), n, p)
```

```{r tied-infer}
res_tied <- infer_pca(
  X_tied, B = 99, R = 10,
  tie_threshold = 0.05, seed = 3
)
res_tied
```

The `Subspace inference (near-tied roots)` line is the headline: one
subspace unit bundles components 1 and 2, and its mean principal angle
across bootstrap replicates is reported directly. The two individual
basis vectors of that plane aren't stable under bootstrap — they rotate
freely — but the plane itself is, and that's what `multifer` measures.

```{r tied-units}
res_tied$units
```

Row 1 is a `subspace` unit with members `c(1L, 2L)` and
`identifiable = FALSE`. Row 2 is the usual `component` unit for the
third, well-separated root.

## Validity contracts fire before the engine runs

Every mature adapter ships executable `checked_assumptions`. Bad input
fails fast with a named reason instead of producing a silent wrong
answer:

```{r validity-na, error = TRUE}
bad <- dat$X
bad[1, 1] <- NA_real_
infer_pca(bad, B = 49, R = 4, seed = 4)
```

```{r validity-const, error = TRUE}
bad <- dat$X
bad[, 3] <- 1.0            # constant column, zero variance
infer_pca(bad, B = 49, R = 4, seed = 5)
```

The checks run at the top of `infer()`, before any SVD or bootstrap. On
clean input the pass records are recorded in `result$assumptions$checked`
for downstream inspection.

## Significance is one question; importance is another

`infer_*()` and the `component_tests` block answer "is component `k`
detectable above noise?" That is a question about **roots**.

A different question — "which variables drive the structure that
ladder selected?" — is answered by `feature_importance_from_bootstrap()`
(descriptive) and `feature_importance_pvalues()` (null-calibrated).
For a worked example see `vignette("pca-inference")` and
`vignette("plsr-inference")`. The schema deliberately separates
significance (per-unit p-values) from importance (per-variable
summaries) so neither is mistaken for the other.

## Where to go next

Five family-specific vignettes walk each inferential family in depth:

- `vignette("pca-inference")` — one-block variance, the collapsed Vitale
  P3 ladder, and variable / score / subspace stability on a signal-bearing
  example.
- `vignette("plsc-inference")` — two-block covariance (PLSC-family), the
  exact core-space bootstrap identity, and cross-domain stability.
- `vignette("cca-inference")` — two-block correlation (CCA-family), the
  mature-tier whitened-space stepwise deflation, and the
  paired-rows / nuisance-adjusted / grouped-exchangeable design matrix.
- `vignette("plsr-inference")` — two-block predictive (PLSR), the
  held-out gain ladder, and feature-importance summaries for the
  selected predictive components.
- `vignette("lda-inference")` — generalized-eigen discriminant-root
  significance under label permutation, plus null-calibrated
  variable importance for class-discriminating features.

For adapter authors, `vignette("writing-adapters")` walks the
contract end to end.

For a quick reference on which adapter ships which family at which
maturity label, see the **Built-in adapters and maturity** table in the
package `README`. `list_infer_adapters(details = TRUE)` returns the
same table directly from the registry.

For the package-level positioning (the
**mature / narrow / planned** taxonomy, the paper-quality v1 roadmap,
and the explicit scope boundary between latent-root families and
predictive / generalized-eigen engines), see `notes/package_vision.md`
in the source tree.
