---
title: "PCA inference with multifer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PCA inference with multifer}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`multifer` provides **typed perturbation inference** for latent-root
multivariate models. This vignette walks the PCA-family case: how to
ask `multifer` a significance / stability question about a one-block
variance decomposition, and how to read the result object.

## What PCA inference looks like in `multifer`

Every call to `infer()` is described by three orthogonal pieces:

- a **geometry** — here `"oneblock"`, meaning "a single data matrix"
- a **relation** — here `"variance"`, meaning "we care about the
  ordered eigenvalues of `X^T X`"
- a **design** — here the default `paired_rows()` is not used; for
  oneblock problems the design is trivial (row exchangeability)

The combination `(oneblock, variance)` is the first **mature-tier**
family in the package. It is paper-backed: the collapsed Vitale P3
tail-root statistic is the observed test statistic, column-permutation
is the null action, and the ladder drives rung-by-rung
stop-at-first-non-rejection.

## Setup

```{r setup}
library(multifer)
```

## A synthetic PCA example with three true components

We build a `40 x 8` matrix with three real signal directions plus
Gaussian noise.

```{r synthetic-data}
set.seed(1)
n <- 40
p <- 8
k <- 3
U <- qr.Q(qr(matrix(rnorm(n * k), n, k)))
V <- qr.Q(qr(matrix(rnorm(p * k), p, k)))
signal <- U %*% diag(c(6, 3, 1.5), nrow = k) %*% t(V)
noise  <- matrix(rnorm(n * p, sd = 0.3), n, p)
X <- signal + noise
```

## Running `infer()`

```{r infer-pca}
res <- infer(
  adapter  = "svd_oneblock",
  data     = X,
  geometry = "oneblock",
  relation = "variance",
  B        = 99,
  R        = 20,
  alpha    = 0.05,
  seed     = 2
)
res
```

The header line counts **units** (latent objects to test) and
distinguishes component units from subspace units. `auto_subspace = TRUE`
is the default, so near-tied roots automatically bundle into a
subspace unit and `print()` leads with the principal-angle stability
summary when one is present.

## Reading the result schema

An `infer_result` is a frozen unit-centered object. The key blocks are:

```{r result-schema}
res$units
res$component_tests
```

`component_tests` carries one row per tested rung with the observed
statistic, the Besag-Clifford sequential p-value, the MC standard
error, and the Phipson-Smyth count `r` of extreme null draws.

Variable and score stability are bootstrap-based stability summaries,
*not* p-values. The schema explicitly separates the two lines (roadmap
point 6 in `notes/package_vision.md`).

```{r stability}
head(res$variable_stability)
head(res$score_stability)
head(res$subspace_stability)
```

## Validity contracts fire before the engine runs

Every mature-tier adapter carries **executable** `checked_assumptions`.
A non-numeric-matrix input, non-finite values, zero-variance columns,
or too-small matrices all fail fast with a named reason.

```{r validity, error = TRUE}
bad <- X
bad[1L, 1L] <- NA_real_
infer(adapter = "svd_oneblock", data = bad,
      geometry = "oneblock", relation = "variance",
      B = 49, R = 4, seed = 3)
```

The checked assumptions are also recorded in the result for the
successful path:

```{r checked-passes}
res$assumptions$checked
```

## What this vignette does *not* cover

- `variable_stability` is a stability measure, not a significance
  test; variable-level significance is deferred to a later phase of
  the package.
- For supervised latent-root inference (PLS regression, reduced-rank
  regression), see the planned `predictive` relation — covariance
  mode is the wrong inferential target for those methods and the
  package refuses to silently reuse it.
