---
title: PCA inference with multifer
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{PCA inference with multifer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 6,
  fig.height = 4,
  dpi       = 150
)
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(family = params$family,
                                              preset = params$preset))
}
```

```{r albers-classes, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family, params$preset
))
```

```{r library}
library(multifer)
```

`multifer` provides **typed perturbation inference** for latent-root
multivariate models. This vignette walks the PCA-family case: how to
ask `multifer` a significance / stability question about a one-block
variance decomposition, and how to read the result object.

## What PCA inference looks like in `multifer`

Every call to `infer()` is described by three orthogonal pieces:

- a **geometry** — here `"oneblock"`, meaning "a single data matrix"
- a **relation** — here `"variance"`, meaning "we care about the
  ordered eigenvalues of `X^T X`"
- a **design** — here the default `paired_rows()` is not used; for
  oneblock problems the design is trivial (row exchangeability)

The combination `(oneblock, variance)` is the first **mature-tier**
family in the package. It is paper-backed: the collapsed Vitale P3
tail-root statistic is the observed test statistic, column-permutation
is the null action, and the ladder drives rung-by-rung
stop-at-first-non-rejection.

## A synthetic PCA example with three true components

We build a `40 x 8` matrix with three real signal directions plus
Gaussian noise.

```{r synthetic-data}
set.seed(1)
n <- 40
p <- 8
k <- 3
U <- qr.Q(qr(matrix(rnorm(n * k), n, k)))
V <- qr.Q(qr(matrix(rnorm(p * k), p, k)))
signal <- U %*% diag(c(6, 3, 1.5), nrow = k) %*% t(V)
noise  <- matrix(rnorm(n * p, sd = 0.3), n, p)
X <- signal + noise
```

## Running `infer_pca()`

`infer_pca()` is the convenience wrapper for the
`(oneblock, variance)` family. It pins the geometry and relation,
picks the `multivarious_pca` adapter by default, and forwards
everything else to `infer()`.

```{r infer-pca}
res <- infer_pca(X, B = 99, R = 20, alpha = 0.05, seed = 2)
res
```

```{r check-pca-result, include = FALSE}
stopifnot(
  is_infer_result(res),
  nrow(res$component_tests) >= 1L,
  sum(res$units$selected) >= 1L,
  all(is.finite(res$component_tests$p_value)),
  all(res$component_tests$p_value >= 0 & res$component_tests$p_value <= 1),
  all(is.na(res$variable_stability$stability) |
        (res$variable_stability$stability >= 0 &
           res$variable_stability$stability <= 1))
)
```

If you need the explicit form — for instance, to swap in a different
adapter while testing — `infer_pca()` is exactly equivalent to
`infer(adapter = "multivarious_pca", data = X, geometry = "oneblock", relation = "variance", ...)`.

## What this vignette tests

```{r pca-labels, echo=FALSE, results='asis'}
cat(knitr::kable(
  data.frame(
    piece = c("Maturity", "Statistic", "Null", "Estimand"),
    value = c(
      "mature",
      res$mc$statistic_label,
      res$mc$null_label,
      res$mc$estimand_label
    ),
    check.names = FALSE
  ),
  col.names = c("piece", "value")
), sep = "\n")
```

The statistic, null, and estimand values above are read from
`res$mc$statistic_label`, `res$mc$null_label`, and
`res$mc$estimand_label`, so this vignette cannot drift from the engine
metadata. `tests/testthat/test-v1-boundary.R` pins this consistency.

The header line counts **units** (latent objects to test) and
distinguishes component units from subspace units. `auto_subspace = TRUE`
is the default, so near-tied roots automatically bundle into a
subspace unit and `print()` leads with the principal-angle stability
summary when one is present.

## Reading the result schema

An `infer_result` is a frozen unit-centered object. The key blocks are:

```{r result-schema}
res$units
res$component_tests
```

`component_tests` carries one row per tested rung with the observed
statistic, the Besag-Clifford sequential p-value, the MC standard
error, and the Phipson-Smyth count `r` of extreme null draws.

Variable and score stability are bootstrap-based stability summaries,
*not* p-values. The schema explicitly separates the two lines (roadmap
point 6 in `notes/package_vision.md`).

```{r stability}
head(res$variable_stability)
head(res$score_stability)
head(res$subspace_stability)
```

## Validity contracts fire before the engine runs

Every mature-tier adapter carries **executable** `checked_assumptions`.
A non-numeric-matrix input, non-finite values, zero-variance columns,
or too-small matrices all fail fast with a named reason.

```{r validity, error = TRUE}
bad <- X
bad[1L, 1L] <- NA_real_
infer(adapter = "svd_oneblock", data = bad,
      geometry = "oneblock", relation = "variance",
      B = 49, R = 4, seed = 3)
```

The checked assumptions are also recorded in the result for the
successful path:

```{r checked-passes}
res$assumptions$checked
```

## Variable importance: the right object for "which variable matters?"

`variable_stability` is a *stability* measure. To ask whether a
variable's loading magnitude is detectable above perturbation or a null,
use the `feature_evidence` sidecar. Bootstrap evidence gives intervals
and bootstrap ratios; p-values are reported only by the null-calibrated
path.

For component-wise evidence, keep the bootstrap artifact and summarize
each feature × selected-unit row. Signed loadings are allowed for
identifiable singleton units. Tied or non-identifiable units
automatically use a subspace norm instead.

```{r feature-evidence-bootstrap}
bundle <- infer_bundle(
  adapter = "multivarious_pca",
  data = X,
  geometry = "oneblock",
  relation = "variance",
  B = 49,
  R = 20,
  seed = 21
)

ev <- feature_evidence_from_bootstrap(
  bundle$artifacts$bootstrap,
  bundle$result$units,
  original_fit = bundle$artifacts$original_fit,
  adapter = bundle$artifacts$adapter,
  statistic = "loading",
  orientation = "auto",
  k = "selected",
  normalize = "none"
)
head(ev[, c("unit_id", "feature", "estimate", "std_error", "z",
            "orientation", "p_value")], 8)
```

```{r check-feature-evidence-bootstrap, include = FALSE}
stopifnot(
  inherits(ev, "multifer_feature_evidence"),
  nrow(ev) > 0L,
  all(ev$calibration == "bootstrap"),
  all(is.na(ev$p_value))
)
```

The `z` column above is a bootstrap ratio, not a p-value. For PCA with
the `multivarious_pca` adapter, method-native conditional subspace
bootstrap is also available through the adapter action:

```{r feature-evidence-native}
native_ev <- feature_evidence_from_adapter(
  adapter = bundle$artifacts$adapter,
  fit = bundle$artifacts$original_fit,
  data = X,
  units = bundle$result$units,
  statistic = "loading",
  orientation = "auto",
  R = 20,
  seed = 22
)
head(native_ev[, c("unit_id", "feature", "estimate", "std_error", "z",
                  "method")], 8)
```

`method = "conditional_subspace_bootstrap"` means the bootstrap is
conditional on the fitted PCA span. It is useful loading evidence, but it
is still not a null-calibrated p-value.

For aggregate p-values, use `feature_evidence_pvalues()`. The statistic
below is an **aggregate** squared-loading statistic over the selected
latent units. With `form_units(adapter$roots(fit))`, every SVD component
is selected, so the question is "does this variable matter across the
fitted component set?"

The setup objects are the same pieces used internally by `infer()`:
`adapter` defines how to fit and read the SVD model, `recipe` defines the
inferential shape, `fit` is the observed SVD fit, and `units` is the
standard table of latent components/subspaces whose loadings are tested.
The `NULL` in `adapter$refit(NULL, X)` is the adapter contract's
`previous_fit` slot; for a fresh fit there is no previous fit.

```{r importance-pvalues}
adapter <- adapter_svd()
recipe <- infer_recipe(
  geometry = "oneblock", relation = "variance",
  adapter  = adapter
)
fit   <- adapter$refit(NULL, X)
units <- form_units(adapter$roots(fit))
fi <- feature_evidence_pvalues(
  recipe, adapter,
  data = X, units = units, original_fit = fit,
  statistic = "squared_loading",
  scope = "aggregate",
  B = 99,  # column-permutation null draws; use more for real analysis
  adjust = "BH", seed = 11
)
head(fi[, c("feature", "statistic", "estimate", "p_value", "p_adjusted")], 8)
```

```{r check-importance-pvalues, include = FALSE}
stopifnot(
  nrow(fi) == ncol(X),
  all(is.finite(fi$estimate)),
  all(fi$p_value >= 0 & fi$p_value <= 1),
  all(fi$p_adjusted >= 0 & fi$p_adjusted <= 1)
)
```

The plot below uses `-log10(p_adjusted)` only to make small adjusted
p-values visually taller. For example, a bar height of `0.08` means
`p_adjusted = 10^-0.08`, about `0.83`; it does **not** mean an adjusted
p-value of `0.08`. The dashed line is the usual `p_adjusted = 0.05`
reference, which appears at `-log10(0.05) = 1.30`.

```{r importance-plot, echo = FALSE, fig.cap = "Evidence for aggregate squared-loading importance. Bar height is -log10(BH-adjusted p-value), so taller bars mean smaller adjusted p-values under the column-permutation null. A height of 1.30 equals adjusted p = 0.05; the dashed line marks that reference level."}
fi$evidence <- -log10(pmax(fi$p_adjusted, .Machine$double.xmin))
ord <- order(fi$evidence, decreasing = TRUE)
fi_o <- fi[ord, ]
col  <- ifelse(fi_o$p_adjusted <= 0.05, "#2ca02c", "#d62728")
threshold <- -log10(0.05)
bp <- barplot(
  fi_o$evidence, names.arg = fi_o$feature, col = col,
  border = NA,
  ylab = expression(-log[10]("BH-adjusted p-value")),
  xlab = "variable",
  main = "aggregate feature-importance evidence, X domain",
  ylim = c(0, max(fi_o$evidence, threshold) * 1.08)
)
abline(h = threshold, lty = 2, col = "gray30")
```

The schema separates significance (per-unit p-values) from importance
(per-variable summaries) on purpose: a stability summary is *not* a
hypothesis test, and the package will not relabel one as the other.

## Cross-references

- For supervised latent-root inference, `vignette("plsr-inference")`
  covers the predictive relation — held-out gain rather than
  in-sample variance.
- For two-block covariance and correlation, see
  `vignette("plsc-inference")` and `vignette("cca-inference")`.
