---
title: PLSC / two-block covariance inference with multifer
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{PLSC / two-block covariance inference with multifer} %\VignetteEngine{knitr::knitr} %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette walks the two-block covariance case: cross-block
inference where the test target is the ordered singular values of
`X^T Y`. This is the **PLSC** (partial least squares, covariance) family
in classical multivariate terminology, and it is the second
mature-tier family in `multifer`.

## Geometry and relation

- geometry: `"cross"` — paired blocks `X` and `Y`
- relation: `"covariance"` — we care about the roots of `X^T Y`
- design: defaults to `paired_rows()` — rows of `X` and `Y` are a
  matched pairing (observation `i` is the same unit in both blocks)

Under the hood, the engine uses an **exact core-space resampling
identity**: given the block thin SVDs
`X = Ux Dx Vx^T` and `Y = Uy Dy Vy^T`, the centered resampled cross
product factors as `Vx M Vy^T` with
`M = Dx · (Ux^T · Uy[perm, ]) · Dy`. The top singular value of the
full product equals the top singular value of the `k_x × k_y` inner
core M, so null draws collapse to a small inner matrix instead of
reforming the full `p × q` cross product at every permutation.

## Setup

```{r setup}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(multifer)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

## A synthetic PLSC example with a shared latent factor

```{r synthetic-cross}
set.seed(1)
n   <- 40
p_x <- 8
p_y <- 6
k   <- 2
U   <- qr.Q(qr(matrix(rnorm(n * k), n, k)))
Wx  <- qr.Q(qr(matrix(rnorm(p_x * k), p_x, k)))
Wy  <- qr.Q(qr(matrix(rnorm(p_y * k), p_y, k)))
sig <- c(5, 2.5)
X <- U %*% diag(sig, nrow = k) %*% t(Wx) +
  matrix(rnorm(n * p_x, sd = 0.5), n, p_x)
Y <- U %*% diag(sig, nrow = k) %*% t(Wy) +
  matrix(rnorm(n * p_y, sd = 0.5), n, p_y)
```

## Running `infer()`

```{r infer-plsc}
res <- infer(
  adapter  = "cross_svd",
  data     = list(X = X, Y = Y),
  geometry = "cross",
  relation = "covariance",
  B        = 99,
  R        = 20,
  alpha    = 0.05,
  seed     = 2
)
res
```

## What this vignette tests

```{r plsc-labels, echo=FALSE, results='asis'}
cat(knitr::kable(
  data.frame(
    piece = c("Maturity", "Statistic", "Null", "Estimand"),
    value = c(
      "mature",
      res$mc$statistic_label,
      res$mc$null_label,
      res$mc$estimand_label
    ),
    check.names = FALSE
  ),
  col.names = c("piece", "value")
), sep = "\n")
```

The statistic, null, and estimand values above come from the result's
engine-provenance labels, not from hardcoded prose.

## Reading the result

```{r cross-units}
res$units
res$component_tests
```

For cross problems, stability summaries are indexed by `(unit_id,
domain)` where `domain` is `"X"` or `"Y"`:

```{r cross-stability}
head(res$variable_stability, 10)
head(res$score_stability, 10)
```

## Subspace stability for tied canonical pairs

When two singular values of `X^T Y` are close, the corresponding
cross-pair is only identifiable up to a 2-dimensional rotation. The
right object to report in that case is the **subspace stability**
(mean and max principal angles between the reference and each
bootstrap fit), which `multifer` exposes directly:

```{r subspace-stability}
head(res$subspace_stability)
```

## Validity contracts

Cross covariance adapters carry the same executable validity
contracts as one-block: numeric matrices, paired rows, finite values,
column variance, and full column rank.

```{r validity, error = TRUE}
bad_X <- X
bad_X[, 2L] <- 1.0  # constant column, zero variance
infer(adapter = "cross_svd", data = list(X = bad_X, Y = Y),
      geometry = "cross", relation = "covariance",
      B = 49, R = 4, seed = 3)
```
