---
title: PLSC / two-block covariance inference with multifer
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{PLSC / two-block covariance inference with multifer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 6,
  fig.height = 4,
  dpi       = 150
)
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(family = params$family,
                                              preset = params$preset))
}
```

```{r albers-classes, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family, params$preset
))
```

```{r library}
library(multifer)
```

You have two matched data blocks, `X` and `Y`, and you want to know
whether their shared covariance structure is stronger than paired-row
noise would produce. This vignette walks that two-block covariance
case. The test target is the ordered singular values of `X^T Y`, the
classical **PLSC** (partial least squares, covariance) family, and it is
the second mature-tier family in `multifer`.

## Geometry and relation

- geometry: `"cross"` — paired blocks `X` and `Y`
- relation: `"covariance"` — we care about the roots of `X^T Y`
- design: defaults to `paired_rows()` — rows of `X` and `Y` are a
  matched pairing (observation `i` is the same unit in both blocks)

## A synthetic PLSC example with a shared latent factor

```{r synthetic-cross}
set.seed(1)
n   <- 40
p_x <- 8
p_y <- 6
k   <- 2
U   <- qr.Q(qr(matrix(rnorm(n * k), n, k)))
Wx  <- qr.Q(qr(matrix(rnorm(p_x * k), p_x, k)))
Wy  <- qr.Q(qr(matrix(rnorm(p_y * k), p_y, k)))
sig <- c(5, 2.5)
X <- U %*% diag(sig, nrow = k) %*% t(Wx) +
  matrix(rnorm(n * p_x, sd = 0.5), n, p_x)
Y <- U %*% diag(sig, nrow = k) %*% t(Wy) +
  matrix(rnorm(n * p_y, sd = 0.5), n, p_y)
```

## Running `infer_plsc()`

`infer_plsc()` is the convenience wrapper. It pins the geometry,
relation, and design, and uses the `multivarious_plsc` adapter by
default. If you want the base-R reference path, request it explicitly
with `adapter = "cross_svd"`:

```{r infer-plsc}
res <- infer_plsc(X, Y, B = 99, R = 20, alpha = 0.05, seed = 2)
res
```

```{r check-plsc-result, include = FALSE}
stopifnot(
  is_infer_result(res),
  nrow(res$component_tests) >= 1L,
  all(is.finite(res$component_tests$p_value)),
  all(res$component_tests$p_value >= 0 & res$component_tests$p_value <= 1),
  any(res$units$selected)
)
```

## What this vignette tests

```{r plsc-labels, echo=FALSE, results='asis'}
cat(knitr::kable(
  data.frame(
    piece = c("Maturity", "Statistic", "Null", "Estimand"),
    value = c(
      "mature",
      res$mc$statistic_label,
      res$mc$null_label,
      res$mc$estimand_label
    ),
    check.names = FALSE
  ),
  col.names = c("piece", "value")
), sep = "\n")
```

The statistic, null, and estimand values above come from the result's
engine-provenance labels, not from hardcoded prose.

Under the hood, the engine uses an **exact core-space resampling
identity**: given the block thin SVDs
`X = Ux Dx Vx^T` and `Y = Uy Dy Vy^T`, the centered resampled cross
product factors as `Vx M Vy^T` with
`M = Dx * (Ux^T * Uy[perm, ]) * Dy`. The top singular value of the
full product equals the top singular value of the `k_x x k_y` inner
core M, so null draws collapse to a small inner matrix instead of
reforming the full `p x q` cross product at every permutation.

## Reading the result

```{r cross-units}
res$units
res$component_tests
```

For cross problems, stability summaries are indexed by `(unit_id,
domain)` where `domain` is `"X"` or `"Y"`:

```{r cross-stability}
head(res$variable_stability, 10)
head(res$score_stability, 10)
```

```{r stability-plot, echo = FALSE, fig.cap = "Variable stability per (unit, domain). Stability of 1 means the variable's loading sign and magnitude are reproduced perfectly across bootstrap replicates; values near 0 mean the loading flips or fluctuates wildly."}
vs <- res$variable_stability
sel <- vs$unit_id %in% res$units$unit_id[res$units$selected]
vs <- vs[sel, , drop = FALSE]
op <- par(mfrow = c(1, 2L), mar = c(4, 4, 3, 1))
tryCatch({
  for (d in c("X", "Y")) {
    vd <- vs[vs$domain == d, , drop = FALSE]
    ord <- order(vd$unit_id, vd$variable)
    vd <- vd[ord, , drop = FALSE]
    col <- ifelse(vd$unit_id == "u1", "#1f77b4", "#ff7f0e")
    barplot(
      vd$stability, names.arg = vd$variable, las = 2L, col = col,
      border = NA, ylim = c(0, 1.05),
      ylab = "stability", main = sprintf("domain %s", d)
    )
  }
}, finally = par(op))
```

Both blocks load mostly on the leading unit, and stability is
high there; the trailing unit recovers signal in only a few
variables, with broader uncertainty.

These stability rows are descriptive bootstrap summaries, not
variable-level p-values. For PLSC/CCA-style cross-decomposition, a
component-wise feature-evidence sidecar should label signed salience
summaries as bootstrap ratios unless a separate null refit calibrates
the statistic. Aggregate squared salience or subspace-norm evidence is
the safer target for feature-level p-values because it is invariant to
sign changes and rotations within tied cross-pair subspaces.

```{r check-plsc-stability, include = FALSE}
selected_vs <- res$variable_stability[
  res$variable_stability$unit_id %in% res$units$unit_id[res$units$selected],
  ,
  drop = FALSE
]
stopifnot(
  nrow(selected_vs) > 0L,
  all(c("X", "Y") %in% selected_vs$domain),
  all(is.finite(selected_vs$stability)),
  all(selected_vs$stability >= 0 & selected_vs$stability <= 1)
)
```

## Subspace stability for tied canonical pairs

When two singular values of `X^T Y` are close, the corresponding
cross-pair is only identifiable up to a 2-dimensional rotation. The
right object to report in that case is the **subspace stability**
(mean and max principal angles between the reference and each
bootstrap fit), which `multifer` exposes directly:

```{r subspace-stability}
head(res$subspace_stability)
```

## Validity contracts

Cross covariance adapters carry the same executable validity
contracts as one-block: numeric matrices, paired rows, finite values,
column variance, and full column rank.

```{r validity, error = TRUE}
bad_X <- X
bad_X[, 2L] <- 1.0  # constant column, zero variance
infer(adapter = "cross_svd", data = list(X = bad_X, Y = Y),
      geometry = "cross", relation = "covariance",
      B = 49, R = 4, seed = 3)
```
