---
title: PLSR / predictive inference with multifer
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{PLSR / predictive inference with multifer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 6,
  fig.height = 4,
  dpi       = 150
)
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(family = params$family,
                                              preset = params$preset))
}
if (!requireNamespace("pls", quietly = TRUE)) knitr::knit_exit()
```

```{r albers-classes-hidden, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family, params$preset
))
```

```{r library}
library(multifer)
```

This vignette covers **supervised predictive inference**. Unlike PLSC,
which tests cross-covariance roots, PLSR inference in `multifer` tests
whether each successive latent predictor adds **held-out predictive
gain**. The ladder is therefore about out-of-sample improvement in `Y`,
not about the singular values of `X^T Y`.

The default public path is `infer_plsr()`, which uses the
`"plsr_refit"` adapter on top of `pls::plsr()`. This public predictive
surface is intentionally narrow in v1: PLSR is shipped, while broader
predictive-cross families remain future work.

## Geometry and relation

- geometry: `"cross"` — a predictor block `X` and a response block `Y`
- relation: `"predictive"` — significance targets held-out predictive gain
- design: defaults to `paired_rows()` — row `i` in `X` predicts row `i` in `Y`

This is a different inferential family from covariance-mode PLSC. The
predictive relation keeps that distinction explicit in the type system
and at adapter registration time.

## A synthetic predictive example

We build two latent predictors in `X`. Both help predict `Y`, but the
first is stronger than the second.

```{r synthetic-plsr}
set.seed(1)
n <- 72

scores <- qr.Q(qr(scale(matrix(rnorm(n * 2L), nrow = n, ncol = 2L),
                         center = TRUE, scale = FALSE)))

x_loadings <- rbind(
  c(1.3, -0.9, 0.8, 0.0, 0.0, 0.0, 0.0, 0.0),
  c(0.0,  0.0, 0.0, 1.1, -0.7, 0.5, 0.0, 0.0)
)
y_loadings <- rbind(
  c(1.2, -1.0, 0.0, 0.0),
  c(0.0,  0.0, 0.9, -0.7)
)

X <- matrix(rnorm(n * 8L, sd = 0.10), nrow = n, ncol = 8L) +
  scores %*% diag(c(3.5, 1.8), nrow = 2L, ncol = 2L) %*% x_loadings
Y <- matrix(rnorm(n * 4L, sd = 0.08), nrow = n, ncol = 4L) +
  scores %*% diag(c(3.2, 1.4), nrow = 2L, ncol = 2L) %*% y_loadings
```

## Running `infer_plsr()`

```{r infer-plsr}
res <- infer_plsr(
  X,
  Y,
  B = 79,
  R = 12,
  alpha = 0.05,
  seed = 2
)
res
```

## What this vignette tests

```{r plsr-labels, echo=FALSE, results='asis'}
cat(knitr::kable(
  data.frame(
    piece = c("Maturity", "Statistic", "Null", "Estimand"),
    value = c(
      "narrow",
      res$mc$statistic_label,
      res$mc$null_label,
      res$mc$estimand_label
    ),
    check.names = FALSE
  ),
  col.names = c("piece", "value")
), sep = "\n")
```

The statistic, null, and estimand values above come from the fitted
result's `mc` provenance labels. They are not hardcoded covariance-root
descriptions.

## Reading the result

The component table reports p-values for successive **incremental
predictive gains**. It is not a cross-covariance-root table.

```{r component-tests}
res$units
res$component_tests
```

The first two latent predictors are selected; the third is not.

## Stability remains a bootstrap layer, not significance

As elsewhere in `multifer`, variable and score summaries are
**stability** outputs. They are useful for reading the fitted latent
predictors, but they are not p-values.

```{r variable-stability}
head(res$variable_stability, 12)
head(res$score_stability, 12)
```

You can already see the expected pattern: variables aligned with the
first predictive direction are more stable than the trailing noise
variables, and the leading score interval is tighter than the trailing
noise unit.

## Variable importance: a descriptive companion

`feature_importance_from_bootstrap()` summarizes the bootstrap
artifact `multifer` already built during `infer_plsr()`'s stability
sweep. Because it consumes the artifact directly, it does **not**
refit anything — it just aggregates squared aligned loadings across
the selected predictive components.

```{r plsr-importance}
adapter <- adapter_plsr_refit()
recipe  <- infer_recipe(
  geometry = "cross", relation = "predictive",
  adapter  = adapter,
  targets  = c("component_significance", "variable_stability",
               "score_stability", "subspace_stability")
)
data    <- list(X = X, Y = Y, relation = "predictive")
fit     <- adapter$refit(NULL, data)
units   <- form_units(adapter$roots(fit))
art     <- bootstrap_fits(
  recipe = recipe, adapter = adapter, data = data,
  original_fit = fit, units = units, R = 12L, seed = 7L
)
fi <- feature_importance_from_bootstrap(
  art, units, k = "all", weights = "root",
  roots = adapter$roots(fit)
)
fi[fi$domain == "X", c("variable", "estimate", "lower", "upper", "stability")]
```

```{r plsr-importance-plot, echo = FALSE, fig.cap = "Aggregate predictive importance per X-variable across the selected latent predictors, with bootstrap 95\\% intervals. The first three variables (the planted predictors of LV1) and the next three (LV2) dominate."}
fx <- fi[fi$domain == "X", , drop = FALSE]
ord <- order(-fx$estimate)
fx <- fx[ord, , drop = FALSE]
op <- par(mar = c(4, 4, 3, 1))
on.exit(par(op), add = TRUE)
mid <- barplot(
  fx$estimate, names.arg = fx$variable, las = 2L,
  col = "#1f77b4", border = NA,
  ylab = "aggregate squared-loading importance",
  main = "feature_importance_from_bootstrap (X domain)",
  ylim = c(0, max(fx$upper, na.rm = TRUE) * 1.05)
)
arrows(mid, fx$lower, mid, fx$upper, code = 3L, angle = 90L,
       length = 0.04, col = "gray30")
```

This is descriptive — not a p-value. For predictive significance at
the *component* level (the held-out gain ladder), see
`res$component_tests` above.

## Why predictive needs its own relation

The predictive family is stricter than covariance-mode PLSC on purpose.
`plsr_refit` is accepted only because it provides:

- `predict_response()`
- a refit path
- a split-aware `component_stat(..., split = NULL)`

That gate prevents in-sample pseudo-significance. For the adapter-side
contract, see `vignette("writing-adapters")`, especially the section
"The predictive relation is stricter on purpose".

## Theory note

For the engine doctrine and the held-out gain definition, see
`notes/engine_predictive_spec.md`.
