---
title: Writing a multifer adapter
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Writing a multifer adapter} %\VignetteEngine{knitr::knitr} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r load}
library(multifer)
```

`multifer` is an **inference layer**, not a fitting framework. If your
package already fits latent-root multivariate models — PCA-like,
covariance two-block, canonical-correlation, or anything that produces
ordered roots plus loadings and scores — you can plug it into
`multifer`'s ladder engine, bootstrap loop, and stability consumers
by writing a thin adapter.

This vignette walks the contract end to end with two complementary
references: a **minimal** base-R wrapper around `base::svd()` (so you
can see every hook spelled out on one screen) and a **realistic**
wrapper around `multivarious::pca()` (so you can see the pattern you
will actually use to adapt a fitted-model class from another package).
The full source for both lives in `R/adapter_oneblock_baser.R` and
`R/adapter_multivarious_pca.R`.

## The contract in one sentence

An adapter tells `multifer` three things:

1. **What** it claims to do (the capability matrix: which targets on
   which (geometry, relation) pairs).
2. **How** to produce the pieces the engine needs from a fitted model
   (the accessor hooks: `roots`, `scores`, `loadings`, `refit`, ...).
3. **What the fit assumes** about the input data (declared and
   executable validity checks).

The engine enforces the first at registration time, dispatches into
the second at inference time, and runs the third before any fit
touches the data.

One practical rule follows from that immediately: if your fitted-model
class is genuinely a single-relation method, declare only that one
relation. `cancor_cross` is a good example. The dual-relation
`cross_svd` adapter is useful as a reference implementation, but its
ambiguity under strict dispatch is intentional and not the default
pattern third-party packages should copy unless they really support
multiple inferential relations from one fit class.

## The minimum surface

Here is a stripped-down version of `adapter_svd()`, the one-block
variance reference. This is the smallest thing that satisfies the
capability gate for every currently shipping target.

```{r minimal-adapter, eval = FALSE}
adapter_minimal_svd <- function() {
  infer_adapter(
    adapter_id      = "minimal_svd",
    shape_kinds     = "oneblock",
    capabilities    = capability_matrix(
      list(geometry = "oneblock", relation = "variance",
           targets  = c("component_significance",
                        "variable_stability",
                        "score_stability",
                        "subspace_stability"))
    ),

    # Accessors -----------------------------------------------------------------
    roots       = function(x, ...) x$d^2,
    scores      = function(x, domain = NULL, ...) x$u %*% diag(x$d),
    loadings    = function(x, domain = NULL, ...) x$v,

    # Deflation for the ladder --------------------------------------------------
    residualize = function(x, k, data, ...) {
      k   <- min(k, length(x$d))
      idx <- seq_len(k)
      recon <- x$u[, idx, drop = FALSE] %*%
               diag(x$d[idx], nrow = k) %*%
               t(x$v[, idx, drop = FALSE])
      data - sweep(recon, 2L, x$center, "+")
    },

    # Perturbation path for the bootstrap ---------------------------------------
    refit = function(x, new_data, ...) {
      if (is.list(new_data) && !is.null(new_data$X)) new_data <- new_data$X
      ctr <- colMeans(new_data)
      fit <- svd(sweep(new_data, 2L, ctr, "-"))
      fit$center <- ctr
      fit
    },

    # Null action for the ladder ------------------------------------------------
    null_action = function(x, data, ...) apply(data, 2L, sample),

    # Observed statistic for the ladder -----------------------------------------
    component_stat = function(x, data, k, ...) {
      s2 <- svd(sweep(data, 2L, colMeans(data), "-"))$d^2
      if (k > length(s2) || sum(s2[k:length(s2)]) == 0) return(NA_real_)
      s2[k] / sum(s2[k:length(s2)])
    },

    # Validity -----------------------------------------------------------------
    validity_level       = "conditional",
    declared_assumptions = c("rows_exchangeable", "centered_data"),
    checked_assumptions  = list()
  )
}
```

That is the whole contract. Every hook has a fixed-arity signature
that the engine calls into; the gate in `infer_adapter()` refuses to
build the adapter if any claimed target is missing its required hook.

### What each hook does, concretely

- `roots(x)` returns the ordered non-negative latent roots. These
  feed [form_units()] and the subspace-detection logic.
- `scores(x, domain)` and `loadings(x, domain)` return the
  per-observation and per-variable latent coordinates. `domain` is
  `"X"` for one-block or `"X"` / `"Y"` for cross-block fits.
- `residualize(x, k, data)` returns `data` with its leading `k`
  components removed. This is the deflation step that moves the
  ladder from rung `k` to rung `k + 1`.
- `refit(x, new_data)` fits a new model of the same class on a
  bootstrap resample. The engine calls this per replicate and then
  aligns the rep to the reference fit.
- `null_action(x, data)` generates one permuted resample under the
  relevant exchangeability assumption. For oneblock variance this
  is a column permutation; for cross-block shapes it breaks the
  pairing by permuting rows of `Y`.
- `component_stat(x, data, k, split = NULL)` returns the observed
  test statistic for rung `k` on `data`. The engine compares this to
  many draws of `component_stat(x, null_action(x, data), k)`. For
  predictive relations, `split` is the held-out fold contract.

## The predictive relation is stricter on purpose

Predictive adapters are the one place where `multifer` refuses the
usual latent-root shortcut. If you claim
`(cross, predictive, component_significance)`, registration-time
validation expects three extra things:

- `predict_response(x, new_data, k = NULL)`, so the engine can score
  held-out responses.
- `refit(x, new_data)`, because predictive significance is evaluated
  by refitting on training folds rather than recycling the original
  fit in-sample.
- `component_stat(x, data, k, split = NULL)` with a formal `split`
  argument. The `split` object carries `$train` and `$test` indices,
  and the hook is expected to compute an out-of-sample gain rather
  than an in-sample covariance or root proxy.

That gate is deliberate. For predictive models such as PLS
regression, an in-sample statistic will rise monotonically with extra
components and badly overstate significance. The contract therefore
builds the reviewer's admissibility rule directly into
`infer_adapter()`: no held-out path, no registration.

## Gates you will hit

Adapter registration is intentionally strict. The most common refusals
are:

- **Predictive cross-fit gate**: an adapter claiming
  `(cross, predictive, component_significance)` must provide
  `predict_response()`, `refit()`, and a `component_stat()` hook with a
  formal `split = NULL` argument. The registration gate lives in
  `R/adapter.R`; the engine contract is described in
  `notes/engine_predictive_spec.md`.
- **Geneig SPD gate**: generalized-eigen operators require a symmetric
  positive-definite metric. The implementation and tests are in
  `R/engine_geneig.R`, `R/linear_operator.R`, and
  `tests/testthat/test-engine-geneig.R`; the design rationale is in
  `notes/engine_geneig_spec.md`.
- **Geneig residualize gate**: adapters claiming
  `(geneig, generalized_eigen, component_significance)` must either mark
  their `residualize` hook as B-metric aware with
  `attr(residualize, "b_metric") <- TRUE` or declare that the geneig
  engine owns deflation with
  `attr(residualize, "delegates_geneig_deflation") <- TRUE`. The
  registration check is in `R/adapter.R`, with the mathematical
  contract in `notes/engine_geneig_spec.md`.

## The realistic wrapper pattern

When your fitted-model class lives in another package — which is
the common case — the adapter becomes a set of delegators into that
package's accessors. Here is the shape of `adapter_multivarious_pca()`:

```{r realistic-adapter, eval = FALSE}
adapter_multivarious_pca <- function() {
  infer_adapter(
    adapter_id      = "multivarious_pca",
    shape_kinds     = "oneblock",
    capabilities    = capability_matrix(
      list(geometry = "oneblock", relation = "variance",
           targets  = c("component_significance",
                        "variable_stability",
                        "score_stability",
                        "subspace_stability"))
    ),

    # Delegate every accessor to multivarious generics --------------------------
    roots    = function(x, ...) x$sdev^2,
    scores   = function(x, domain = NULL, ...) multivarious::scores(x),
    loadings = function(x, domain = NULL, ...) multivarious::components(x),

    # Deflation via multivarious's own truncation + reconstruction -------------
    residualize = function(x, k, data, ...) {
      trunc <- multivarious::truncate(x, k)
      recon <- multivarious::reconstruct(trunc)
      data - recon
    },

    # Refit by calling multivarious::pca() on the resampled data ---------------
    refit = function(x, new_data, ...) {
      if (is.list(new_data) && !is.null(new_data$X)) new_data <- new_data$X
      multivarious::pca(new_data, ncomp = length(x$sdev))
    },

    null_action = function(x, data, ...) apply(data, 2L, sample),

    component_stat = function(x, data, k, ...) {
      fit <- multivarious::pca(data, ncomp = k)
      s2  <- fit$sdev^2
      if (k > length(s2) || sum(s2[k:length(s2)]) == 0) return(NA_real_)
      s2[k] / sum(s2[k:length(s2)])
    },

    validity_level       = "conditional",
    declared_assumptions = c("rows_exchangeable", "centered_data"),
    checked_assumptions  = list()
  )
}
```

The shape is the same; only the bodies change. Each hook is a thin
delegator to a `multivarious::` generic — no math, just plumbing.
That is what we mean by "multifer is the inference layer on top of
a fitting framework." Your package provides the fitter; `multifer`
provides the ladder.

## Registering the adapter

Adapters are looked up by `adapter_id` through a per-session
registry. Call [register_infer_adapter()] once, typically from your
package's `.onLoad` hook:

```{r register, eval = FALSE}
.onLoad <- function(libname, pkgname) {
  multifer::register_infer_adapter(
    adapter_id = "minimal_svd",
    adapter    = adapter_minimal_svd(),
    overwrite  = TRUE
  )
}
```

Once registered, users call it by id:

```{r use-registered, eval = FALSE}
infer(
  adapter  = "minimal_svd",
  data     = X,
  geometry = "oneblock",
  relation = "variance"
)
```

or by passing the adapter object directly:

```{r use-direct, eval = FALSE}
infer(
  adapter  = adapter_minimal_svd(),
  data     = X,
  geometry = "oneblock",
  relation = "variance"
)
```

## Validity contracts

The `checked_assumptions` slot is the executable safety layer. Every
entry is `list(name, check, detail)`. [infer()] runs each `check`
against the raw data before the engine starts:

```{r validity, eval = FALSE}
.my_checks <- function() {
  list(
    list(
      name   = "data_is_numeric_matrix",
      detail = "`data` must be a numeric matrix",
      check  = function(data, ...) is.matrix(data) && is.numeric(data)
    ),
    list(
      name   = "data_is_finite",
      detail = "`data` must not contain NA, NaN, or Inf",
      check  = function(data, ...) all(is.finite(data))
    )
  )
}
```

If any check fails in strict mode (the default), `infer()` errors
with the violation's `detail` message before any SVD runs. In
permissive mode (`strict = FALSE`), violations are recorded in
`result$assumptions$checked` and the engine proceeds.

The reference adapters ship helpers you can reuse directly:

```{r helpers, eval = FALSE}
# Reuse the base-R adapter checks (not exported; see R/adapter_*.R)
adapter_minimal_svd <- function() {
  infer_adapter(
    # ... other fields ...
    checked_assumptions = .oneblock_baser_checks()
  )
}
```

These cover the common failure modes — non-numeric matrix, non-finite
entries, zero-variance columns, rank deficiency — so third-party
adapters for oneblock variance or cross families can inherit them
without rewriting.

## The Phase 1.5 fast path (optional)

If your fitted class admits a **core representation** with a cheap
resample update — for example, a thin SVD whose column spaces are
orthonormal — you can add two more hooks:

```{r fast-path, eval = FALSE}
infer_adapter(
  # ... other fields ...
  core        = function(x, data, ...)      { ... },
  update_core = function(core_obj, indices = NULL, ...) { ... }
)
```

When both hooks are present, the bootstrap loop uses `update_core()`
instead of `refit()`. For one-block variance this turns `n × p^2`
refits into `n × k^2` inner SVDs (the Fisher identity); for
covariance-mode cross it collapses to a `k_x × k_y` inner core via
the identity `X^T W Y = V_x · (D_x · U_x^T W U_y · D_y) · V_y^T`.
See `R/adapter_oneblock_baser.R` and `R/core_space.R` for the
worked examples.

The core hooks are strictly optional. The engine always falls back
to `refit()` when the fast path is unavailable, so you can ship a
correct adapter first and add the fast path later as an
optimization.

## Testing your adapter

A minimum integration test calls [infer()] with your adapter and
verifies the result object looks right:

```{r test-stub, eval = FALSE}
test_that("my_adapter runs end to end", {
  X <- matrix(rnorm(40 * 6), 40, 6)
  res <- infer(
    adapter  = adapter_minimal_svd(),
    data     = X,
    geometry = "oneblock",
    relation = "variance",
    B = 49L, R = 6L, seed = 1L
  )
  expect_true(is_infer_result(res))
  expect_true(nrow(res$component_tests) >= 1L)
  expect_true(nrow(res$variable_stability) >= 1L)
})
```

For correctness, compare your adapter's fast path to its `refit`
path across a scenario grid (see `tests/testthat/test-core-update-*.R`
for the pattern we use internally). The package also ships
`tests/testthat/test-validity-checks.R` as a template for asserting
that your `checked_assumptions` fire on the failure modes you care
about.

## Where to go next

- `?infer_adapter_contract` — the canonical contract reference with
  every hook signature and every capability-gate rule.
- `?infer_adapter` — the constructor itself.
- `R/adapter_oneblock_baser.R` — the full `adapter_svd()` and
  `adapter_prcomp()` reference sources.
- `R/adapter_multivarious_pca.R` — the fitted-model-wrapper pattern
  in ~100 lines.
- `vignette("multifer")` — end-to-end inference usage from the
  user's side, once you have registered an adapter.
